Interactive Checks for Coordination Avoidance

### 2019-08-16 doc/vldb_slides/sections/bank_example.tex
Started writing speaker notes.

\begin{frame}{Replicated Bank Account}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance, 0) {\$50};
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$+\$10$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$+\$20$};
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$-\$30$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$-\$40$};
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}
  \begin{center}
    \Huge
    Deposits don't require coordination to maintain invariants, but withdrawals
    do.
  \end{center}
\end{frame}

>>>>

\begin{frame}{Replicated Bank Account}
  \begin{center}
    \begin{tikzpicture}
      \node[bank, label=above:{Replica $A$}] (a1) at (0, 0) {\$50};
      \node[bank, label=above:{Replica $B$}] (b1) at (\bankdistance, 0) {\$50};
    \end{tikzpicture}
  \end{center}

  \note{%
    So, what is invariant confluence? Well, let's begin with an example.

    Let's say we want to replicate the database shown here. This database is
    the world's simplest database. It contains only a single number, which
    represents my bank account balance. We decide to replicate this database
    across two machines, replica A and replica B.

    Users can issue transactions to the database, and every transacation is
    either a deposit or a withdrawal. And, I have one critical invariant that I
    want to maintain over my database. And that invariant is that my bank
    account balance should be non-negative at all times. I'm not allowed to go
    in the hole.
  }
\end{frame}

@@

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}

  \note{%
    One way to replicate our database is by using coordination to implement
    strong consistency. For example, imagine a user issues a deposit of 10
    dollars to replica A, and simultaneously, a user issues a deposit of 20
    dollars to replica B. The two replicas can coordinate, for example by
    running Paxos, and agree on a single total order in which to execute the
    two transactions. Here, the replicas agree to deposit 10 dollars and then
    deposit 20 dollars. Both transactions succeed, and the replicas send back OK
    messages to the clients letting them know everything went smoothly.
  }
\end{frame}

@@

      \zigzag{a1}{b1}
      \zigzag{a2}{b2}
    \end{tikzpicture}
  \end{center}

  \note{%
    Now, assume two clients issue concurrent withdrawal transactions of 30
    dollars and 40 dollars. Again, the two replicas coordinate and agree to
    execute the withdrawal of 30 dollars and then the withdrawal of 40 dollars.

    The withdrawal of 30 dollars executes fine, brining my bank account balance
    to 20 dollars, but the withdrawal of 40 dollars fails because I only have
    20 dollars left. So replica A sends back an OK message to the client, but
    replica B sends back a NO message letting the client know the transaction
    failed. And while the transaction failed, you'll note that we never
    violated our application invariant.
  }
\end{frame}

@@

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$+\$10$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$+\$20$};
    \end{tikzpicture}
  \end{center}

  \note{%
    Now, instead of implementing our bank with strong consistency, we could
    instead implement it with weak consistency and avoid coordination
    altogether.

    Now, when two clients issue concurrent deposit transactions, the two
    replicas execute the transactions immediately without consulting one
    another. And every once in a while, the two replicas communicate and fill
    each other in on the transactions they missed.

    Here, for example, replica A lets replica B know to deposit 10 dollars and
    replica B lets replica A know to deposit 20 dollars. Notice that even
    though we didn't perform any coordination, the two replicas never violate
    my critical invariant of a non-negative balance.
  }
\end{frame}

@@

      \draw[dashed, op] (a2) -- (b3) node[near start, sloped, above] {$-\$30$};
      \draw[dashed, op] (b2) -- (a3) node[near start, sloped, above] {$-\$40$};
    \end{tikzpicture}
  \end{center}

  \note{%
    However, let's look what happens when two clients issue concurrent
    withdrawals. Both withdrawals are executed locally and both succeed.
    However, when the two replicas exchange transactions with one another, they
    both enter an invariant violating state of negative twenty dollars!
  }
\end{frame}

\begin{frame}
  \begin{center}
    \Huge
    Deposits don't require coordination to maintain invariants, but withdrawals
    do.
  \end{center}

  \note{%
    Now, these are just two examples, but if you think about it long enough,
    they are representative of a trend. We can implement deposit transactions
    without any coordination and we're guaranteed to never violate the
    invariant. Withdrawal transactions, on the other hand, do require
    coordination to avoid violating the invariant.

    At a very high level, this is exactly the notion that invariant confluence
    captures. We'll see that deposit transactions are invariant confluent and
    thus do not require coordination while withdrawal transactions are not
    invariant confluent and thus do require coordination.
  }
\end{frame}

### 2019-08-16 doc/vldb_slides/sections/four_wise_men.tex
Started writing speaker notes.

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{You should use weak consistency. It's wicked fast!}
      };
      \path[fill=gray!30] (vogels.north) -- ++(0, 1) -- ++(0.5, 0) -- (vogels.north);
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \node[quote, text width=0.8\textwidth, align=left] at (1, 3) {
        \Large\textit{What about the application invariants?!}
      };
      \path[fill=gray!30] (hoare.north) -- ++(0, 1.5) -- ++(1, 0) -- (hoare.north);
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \node[quote, text width=0.8\textwidth, align=left] at (2, 3) {
        \Large\textit{Yeah, go with strong consistency! I have an algorithm for that :)}
      };
      \path[fill=gray!30] (lamport.north) -- ++(0, 1.5) -- ++(1, 0) -- (lamport.north);
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \brewer
      \node[quote, text width=0.8\textwidth, align=left] at (3, 3) {
        \Large\textit{Not so fast!}
      };
      \path[fill=gray!30] (brewer.north) -- ++(0, 1.5) -- ++(-1, 0) -- (brewer.north);
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {\includegraphics[width=0.25\textwidth]{assets/bailis.jpg}};
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{?`Porque no los dos?}
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}
\end{frame}

>>>>

\begin{frame}
  \note{%
    Hi everyone! Thanks so much for attending. I'm Michael, and today I'll be
    talking about ``Interactive Checks for Coordination Avoidance''. This is
    joint work with my advisor at UC Berkeley, Joe Hellerstein.

    Imagine you're an application developer and you're trying to decide what
    kind of consistency you'd like to use to replicate some of your data.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{You should use weak consistency. It's wicked fast!}
      };
      \path[fill=gray!30] (vogels.north) -- ++(0, 1) -- ++(0.5, 0) -- (vogels.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Someone like Werner Vogels would probably tell you to use weak consistency.
    You can implement weak consistency super duper fast, and for many
    applications, weak consistency is good enough.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \node[quote, text width=0.8\textwidth, align=left] at (1, 3) {
        \Large\textit{What about the application invariants?!}
      };
      \path[fill=gray!30] (hoare.north) -- ++(0, 1.5) -- ++(1, 0) -- (hoare.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    But then Tony Hoare laments, ``no no no''. Hoare encourages us not to
    settle for weak consistency because a weakly consistent system is hard to
    reason about. It robs us of a fundamental way we use to reason about
    programs: application invariants. Even if every transaction in a weakly
    consistent system preserves application invariants, the whole system may
    not.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \node[quote, text width=0.8\textwidth, align=left] at (2, 3) {
        \Large\textit{Yeah, go with strong consistency! I have an algorithm for that :)}
      };
      \path[fill=gray!30] (lamport.north) -- ++(0, 1.5) -- ++(1, 0) -- (lamport.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Leslie Lamport chimes in agreeingly. He encourages us to choose strong
    consistency. Strongly consistent systems are much easier to reason about.
    After all a strongly consistent system to end users looks exactly like a
    system that's not even distributed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \brewer
      \node[quote, text width=0.8\textwidth, align=left] at (3, 3) {
        \Large\textit{Not so fast!}
      };
      \path[fill=gray!30] (brewer.north) -- ++(0, 1.5) -- ++(-1, 0) -- (brewer.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    ``Aha, not so fast'' counters Eric Brewer! Consistency and availability and
    partition tolerance, oh my. Surely, strong consistency comes at the price
    of availability and performance.

    So we're in a bit of a pickle. Weak consistency is too confusing and strong
    consistency is too heavy handed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {\includegraphics[width=0.25\textwidth]{assets/bailis.jpg}};
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{?`Porque no los dos?}
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Thankfully, Peter Bailis comes to save the day and reminds us that perhaps
    we can have the best of both worlds. In 2014, Peter and some other smart
    folks defined invariant confluence as a condition under which we can
    replicate a piece of data without the need for strongly consistent
    coordination while maintaining the data's application invariants.

    Invariant confluence will be the topic of this talk. In particular, I'll
    define invariant confluence and then discuss how we can check invariant
    confluence automatically.
  }
\end{frame}
